\documentclass[doublecol]{epl2}

\usepackage{amsmath, amssymb}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage{hyperref}

\title{Comment on ``Measurements of Newton's gravitational constant and the length of day''}

\author{M. Pitkin}

\institute{
  SUPA, School of Physics \& Astronomy, University of Glasgow, Glasgow, G12 8QQ, UK
}

\pacs{04.80.-y}{Experimental studies of gravity}
\pacs{07.05.Kf}{Data analysis: algorithms and implementation; data management}

\begin{document}

\maketitle

\section{Introduction}

In \cite{2015EL....11010002A} the authors claim to observe a periodic signal in measurements of
Newton's gravitational constant, $G$. Specifically they find a signal with a period of 5.9\,years that
is very strongly correlated with variations in the observed length of day
\cite{2013Natur.499..202H}. Here I present a reanalysis of the data used in \cite{2015EL....11010002A}
performing Bayesian model selection to test the significance of the hypothesis that the data does contain
a periodic signal compared to other potential models. In light of the updated information on the times
of the various $G$ measurements given in \cite{2015arXiv150501774S}, which have been shown to considerably
reduce the significance of the claim in \cite{2015EL....11010002A}, I also reanalyse this new dataset with
the same method. In both datasets I have found that a model for the variations that only contains an
additional Gaussian noise term is hugely favoured, by factors of $\gtrsim e^{30}$, over models containing
an additional sinusoid term.\footnote{The code, data tables, figures and prior ranges for this analysis can be
found at \url{https://github.com/mattpitkin/periodicG}.}

\section{Analysis method}

Bayesian model selection provides a natural way to test multiple hypotheses by forming the Bayesian odds ratio of
evidences for the different hypotheses. The Bayesian odds ratio for two hypotheses $H_i$ and $H_j$ is given by
\begin{equation}
 \mathcal{O}_{ij} = \frac{p(d|H_i,I)}{p(d|H_i,I)}\frac{p(H_i|I)}{p(H_j|I)}
\end{equation}
where $p(d|H_i,I)$ is the evidence (or marginal likelihood) for hypothesis $H_i$ given some data $d$,
$p(H_i|I)$ is the prior probability for $H_i$, and $I$ is information concerning any other assumptions used to
define the problem at hand. When comparing hypotheses I assume that they are equally
probable a priori, so the prior ratio is unity. Therefore, I just calculate
the ratio of evidences for each hypothesis (often called the Bayes factor). If a given hypothesis is defined
by a model containing a set of parameters, $\vect{\theta}_i$, with their own priors,
$p(\vect{\theta}_i|H_i,I)$, then to calculate the evidence, $Z_i$, the parameters must be marginalised (i.e.\ integrated)
over, e.g.\
\begin{equation}\label{eq:evidence}
 p(d|H_i,I) \equiv Z_i = \int^{\vect{\theta}} p(d|\vect{\theta}_i,H_i,I) p(\vect{\theta}_i|H_i,I) {\rm d}\vect{\theta}_i,
\end{equation}
where $p(d|\vect{\theta}_i,H_i,I)$ is the likelihood function of the data given at set of model parameters $\vect{\theta}_i$.
In the hypotheses I define below the number of parameters are small enough that the likelihoods can be evaluated on a grid in the parameter space and the integrals performed numerically using the trapezium rule.

The general model that I use for my hypotheses includes a sinusoid and an offset
\begin{equation}\label{eq:model}
 m(\mu_G, A, P, \phi_0, T_k) = A\sin{(\phi_0 + 2\pi (T_k-t_0)/P)} + \mu_G,
\end{equation}
where $\mu_G$ is the offset value, $A$ is the sinusoid amplitude, $\phi_0$ is an initial phase at an epoch $t_0$,
$P$ is the sinusoid period, and $T_k$ is the time.

In this analysis I have tested four different hypotheses of increasing complexity to explain the measurements of $G$:
\begin{enumerate}
 \item the data is consistent with Gaussian errors, given by the experimental error bars, about an unknown
 offset value,
 \item the data is consistent with Gaussian errors, given by the experimental error bars, about an unknown
 offset value {\it and} an unknown common Gaussian noise term,
 \item the data is consistent with Gaussian errors, given by the experimental error bars, about an unknown
 offset value and a sinusoid with unknown amplitude, initial phase and period,
 \item the data is consistent with Gaussian errors, given by the experimental error bars, about an unknown
 offset value, an unknown common Gaussian noise term and a sinusoid with unknown amplitude, initial phase and period.
\end{enumerate}
These each correspond to a different set of parameters required in $\vect{\theta}$ and also the number of
parameter required in the integral of eq.~\ref{eq:evidence}.

For an initial examination of the claim in \cite{2015EL....11010002A} I have used their Figure~1 to read-off
the experimental times and then (generally) used Table~XVII of \cite{RevModPhys.84.1527} for values of $G$.
For the BIPM-13 measurements I used the combined servo and Cavendish value from \cite{PhysRevLett.113.039901}
and for the LENS-14 measurements I used the values from \cite{2014Natur.510..518R}.

In \cite{2015EL....11010002A} the given experiment times are assumed to be correct with no associated error.
However, many of the times used correspond to the received date of the respective paper rather than the
date of the actual experiment.  In analysing this data I specified uncertainties on the experiment times of
$\sigma_{t,i} = 0.25$\,years (with the exception of the JILA-10 and LENS-14 measurements for which I use uncertainties
of one week) before the given time. I have taken into account this uncertainty on the
time of each data point, by marginalising over the uncertainty for each data point.

\section{Results}

The odds ratios comparing hypotheses when using the $G$ dataset of \cite{2015EL....11010002A} are summarised in
table~\ref{tab:results}. It is clear that hypotheses including extra parameters over that for hypothesis 1
(i.e.\ including a sinusoidal component and/or and extra unknown noise component) are hugely favoured by factors
of $\gtrsim e^{100}$. The two hypotheses, $H_3$
and $H_4$, containing a sinusoidal signal are both approximately equally probable. However, hypothesis 2, just
containing the additional unknown noise term and the unknown offset, is hugely favoured by factors $\sim e^{30}$
over $H_3$ and $H_4$. This shows that the simple model that variations are just due to an unknown Gaussian
noise term is far more likely to be the cause of the variations than an additional sinusoidal variation.
This is due to Bayesian model selection naturally applying a penalty for including additional parameters that
do not significantly increase the evidence.

\begin{table}
\caption{Bayesian odds ratios for the four hypotheses ($i$ represents rows and $j$ represents columns)
when using the data used in \cite{2015EL....11010002A}.}
\label{tab:results}
\begin{center}
 \begin{tabular}{c|ccc}
 & \multicolumn{3}{c}{$\mathcal{O}_{ij} = Z_i/Z_j$} \\
 \hline
   & $Z_2$ & $Z_3$ & $Z_4$ \\
  \specialrule{0.25pt}{0.75pt}{0.75pt}
  $Z_1$ & $e^{-132.5}$ & $e^{-102.2}$ & $e^{-102.5}$ \\ 
  $Z_2$ &  & $e^{30.3}$ & $e^{29.9}$ \\
  $Z_3$ &  &  & $0.74$
 \end{tabular}
\end{center}
\end{table}

It is, however, interesting to look at the posterior probability distributions for each of the parameters used
in hypotheses. From the posteriors on the period for hypothesis 3 I see a clear large spike in probability
around the claimed period of 5.9\,years. A similar spike shows up in for hypothesis 4, but is much less pronounced.
However, I have also assessed the significance of the period peak for hypothesis 3 by rerunning that analysis 20 times,
but each time randomly shuffling the $G$ values (whilst keeping the measurement times the same). This should remove
any real periodicity in the data.  Out of these 20 runs there was one time when the hypothesis using the shuffled data
is more favoured than when using the un-shuffled data and another couple within a factor of two. The posteriors for
these case also show that very similar spikes in the period can be produced in randomised data.

\section{Updated dataset}

Following my initial investigations of the claims of \cite{2015EL....11010002A} Schlamminger, Gundlach and Newman
also examined the claim \cite{2015arXiv150501774S} in particular noting that the experimental times (and also number 
of experimental data used) in the original work are not accurate. They examined the literature to compile a more complete list of
experiments with information on the actual dates that the experiments were
performed.

I have reanalysed this new dataset for each of the four hypotheses. When integrating over the
time error I have now set the error to be symmetric around the mean experiment time.
For all other parameters I have used the same prior ranges as before.
The Bayesian odds ratios for each of these cases are given in table~\ref{tab:newresults} from which it can
be seen that hypothesis 2 is still favoured over all other hypotheses by a huge amount. However, hypothesis 3
is now hugely disfavoured over hypothesis 4, i.e.\ just including a sinusoid, but adding no additional noise
term does far worse at fitting the data than also including the noise term.

\begin{table}
\caption{Bayesian odds ratios for the four hypotheses ($i$ represents rows and $j$ represents columns)
when using the data from \cite{2015arXiv150501774S}.}
\label{tab:newresults}
\begin{center}
 \begin{tabular}{l|ccc}
 & \multicolumn{3}{c}{$\mathcal{O}_{ij} = Z_i/Z_j$} \\
 \hline
   & $Z_2$ & $Z_3$ & $Z_4$ \\
  \specialrule{0.25pt}{0.75pt}{0.75pt}
  $Z_1$ & $e^{-140.2}$ & $e^{-65.8}$ & $e^{-110.3}$ \\
  $Z_2$ & & $e^{74.4}$ & $e^{29.9}$  \\
  $Z_3$ & & & $e^{-44.5}$
 \end{tabular}
\end{center}
\end{table}

\section{Conclusions}

I have reanalysed the data consisting of measurements of Newton's gravitational constant $G$ from
\cite{2015EL....11010002A} and \cite{2015arXiv150501774S} to asses the claim a periodic
component with a period of 5.9\,years. It was not suggested that $G$ was actually varying on these
timescales, but rather that there could be some systematic effect on the measurement process that was
correlated with the mechanism that leads to the variation in the length of the day.

Using Bayesian model selection, and four different hypotheses to describe the variations in the data, and including
uncertainties on the experimental times, I have found that the best model to describe the data is one in which
there is an additional common unknown Gaussian noise term on top of the oberserved experimental values. This is 
favoured over a model also containing a sinusoidal term by factors of $\gtrsim e^{30}$. I also find that periodic
signals can easily be found in random permutations of the data suggesting that the observed periodicity seen in
\cite{2015EL....11010002A} is just a random artifact of the data.

I note that if there were very good a priori reasons to expect a periodic component in the data, and a known
period range and/or initial phase (i.e.\ if there were a good reason why the the mechanism leading to changes in the
length of the day could couple into measurements of $G$), then the evidence in favour models containing a periodic
signal would dramatically increase. However, without such prior knowledge using such a constraint would strongly bias
us.

Since starting this work the authors of the original work have added an Appendix (see \cite{AndersonRevised}) taking
into account the work of \cite{2015arXiv150501774S}. They fit an additional sinusoid to the data and note that the
significance of the correlation with the length of day decreases. I expect that calculating the evidence for a model
including two sinusoids would not lead to such a model being favoured over the simpler model containing just the
extra Gaussian noise term, due to the increase in parameter space being penalised if they fit does not significantly
improve.

\acknowledgements

I would like to thank Prof.\ J.\ Faller for useful discussions, Prof.\ C.\ Speake for putting me in contact with
Dr.\ S.\ Schlamminger, and Dr.\ Schlamminger for providing me with a data file of their compiled $G$ measurements.
I am funded by the STFC under grant ST/L000946/1.

\bibliographystyle{eplbib}
\bibliography{comment_revised}

\end{document}
